---
title: "merus length vs carapace width: exploring Bristol Bay data"
author: "Katie Palof  - consult on project for L.Hulbert"
date: "March 21, 2018"
output: word_document
---

Data presented here is taken from Donaldson & Blackburn 1989. THe objective of their study was to "describe the merus length - carapace width relationship for male red king crab of the Bristol Bay population". 

Caveat: In this exploratory analysis we are using Bristol Bay data but the legal size for Southeast Alaska. One would expect the data maybe different for Southeast Alaska due to difference in growth. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
options(scipen=999)
library(xlsx)
library(extrafont)
library(pwr)
loadfonts(device="win")
windowsFonts(Times=windowsFont("TT Times New Roman"))

theme_set(theme_bw(base_size=12,base_family='Times New Roman')+ 
            theme(panel.grid.major = element_blank(),
                  panel.grid.minor = element_blank()))

dat <- read.xlsx("./data/Donaldson Blackburn 1989 Data.xlsx", sheetName = "rinput")
```

## Linear relationship between carapace width and raw merus length


```{r linear, echo = FALSE}
ggplot(dat, aes(width, raw_merus)) + 
         geom_point(size =2) +
         geom_smooth(method =lm)

fit1 = lm(raw_merus ~ width, data = dat) 
summary(fit1)
#plot(fit1)
```

## Confidence and Prediction intervals

```{r intervals, echo=FALSE}
# confidence interval -----
new.dat <- data.frame(width = seq(from =135, to = 215, length.out = 181))
conf_inter <- predict(fit1, newdata = new.dat, interval = 'confidence')
conf_inter <- as.data.frame(conf_inter)

conf_inter %>% 
  bind_cols(new.dat) ->confidenceI

# prediction interval ------
pred_inter <- predict(fit1, newdata = new.dat, interval = 'prediction')
pred_inter <- as.data.frame(pred_inter)

pred_inter %>% 
  bind_cols(new.dat) ->predictionI


# visual of confidence and predictive intervals -----
ggplot(dat, aes(width, raw_merus)) +
        geom_point(size =2, pch = 1) +
        geom_line(data = confidenceI, aes(width, fit), color = "black", size = 1) +
        geom_line(data = confidenceI, aes(width, lwr), color = "red", lty = 2, size = 1) +
        geom_line(data = confidenceI, aes(width, upr), color = "red", lty =2, size = 1) +
        geom_line(data = predictionI, aes(width, lwr), color = "blue", lty = 3, size = 1) +
        geom_line(data = predictionI, aes(width, upr), color = "blue", lty =3, size = 1) +
  xlab("carapace width (mm)") + ylab("merus length (mm)") + ggtitle("carapace width vs raw merus length") +
  geom_vline( xintercept = 177.8)+
  geom_hline(yintercept = 136.7) +
  geom_vline(xintercept = 165, color = "grey41", lty = 2) +
  geom_hline(yintercept = 126.7, color = "grey41", lty = 2)
```


Footnote: 

Black line is linear regression, red lines are confidence intervals and blue dashed lines are the prediction intervals.  Vertical and horisontal black line represents legal size in SE Alaska. Vertical and horizontal dashed gray line represents legal size in Bristol Bay (where data are taken from). 

\newline


## Goal: sample size needed to reduce variability in the regression line and therefore decrease the 95% prediction interval 

Not sure if this is possible, increasing sample size will strengthen the linear relationship but will not reduce the underlying variability in this relationship. Suggestions for data sampling include:

1) Sampling among carapace width bins to ensure coverage of the entire range of data, up to researchers to determine best bin widths.   
2) Sample more intensely in bins that are within the area where there is overlap (see rectangle in figure below).  

\newline

3) Consider factors that affect growth and see if we can use these as covariates in this relationship.  Example: shell condition?   

4) Spatial considerations on this relationship.


\newline

\newline

``` {r sample area, echo=FALSE}
ggplot(dat, aes(width, raw_merus)) +
        geom_point(size =2, pch = 1) +
        geom_line(data = confidenceI, aes(width, fit), color = "black", size = 1) +
        #geom_line(data = confidenceI, aes(width, lwr), color = "red", lty = 2, size = 1) +
        #geom_line(data = confidenceI, aes(width, upr), color = "red", lty =2, size = 1) +
        geom_line(data = predictionI, aes(width, lwr), color = "blue", lty = 3, size = 1) +
        geom_line(data = predictionI, aes(width, upr), color = "blue", lty =3, size = 1) +
  xlab("carapace width (mm)") + ylab("merus length (mm)") + ggtitle("Intense sample area: carapace width vs raw merus length") +
  geom_vline( xintercept = 177.8, color = "red") + 
  #geom_hline(yintercept = 125, color = "dark grey", size = 0.8) +
  #geom_hline(yintercept = 144, color = "dark grey", size = 0.8) +
  #geom_vline(xintercept = 151, color = "dark grey", size = 0.8) +
  #geom_vline(xintercept = 194, color = "dark grey", size = 0.8) +
  scale_x_continuous() +
  scale_y_continuous() +
  geom_rect(data = dat, mapping = aes(xmin =151, xmax = 194, ymin= 125, ymax = 144), color= "gray52", alpha =0.0005)

```

The carapace width range in this box is:
```{r rectangle, echo= FALSE}
# seperate data into legal and non groups
dat %>% 
  mutate(legal = ifelse(width < 178, "N", "Y")) -> dat
dat %>% 
  filter(raw_merus >124 & legal =="N") %>% 
  summarise(min = min(width))

dat %>% 
  filter(legal == "Y" & raw_merus <144) %>% 
  summarise(max = max(width))
```

\newline

\newline
\newline

## Power analysis 

### sample size for linear regression
The general guidline for a linear regression with a significance level of .05 and a power of 85% without prior knowledge would suggest a sample size of 449 (see results below [1]). Increasing the power to 95%, under the same conditions would yield a same size of 650.

\newline

However, using the knowledge of the existing relationship above will decrease the sample size needed by increasing the effect size (because you already know there is a strong linear relationship - adjusted $R^2$ = 0.7471). This suggests a sample size of 150 (alpha = 0.05, power = 85%) or a sample size of 217 (alpha = 0.05, power = 95%), depending on the desired power.


```{r power1, echo=FALSE}
## power analysis ------------
# power analysis to determine sample size needed for a linear regression.
# use f2 suggested be cohen, small, medium, and large (0.05=2,0.15,0.35)
print("General regression sample size")
pwr.f2.test(u=1, f2 =0.02, sig.level = 0.05, power = 0.85)
#pwr.f2.test(u=1, f2 =0.02, sig.level = 0.05, power = 0.95)

print("Using knowledge of the relationship decreases neccessary sample size")
pwr.f2.test(u=1, f2 =0.06, sig.level = 0.05, power = 0.85)
pwr.f2.test(u=1, f2 =0.06, sig.level = 0.05, power = 0.95)
```


### Power to detect difference between mean merus length of legal and sublegal?
Conclusion: already a large difference between these two means power analyis for this will NOT help to reduce variablity in the regression and therefore the prediction interval.

#### Summary of all data: Mean carapace width and raw merus length
```{r power2, echo=FALSE}
## power analysis ------------
# power analysis to determine sample size needed to predict if legal or not.

#average merus length overall
dat %>% 
  summarise(avg_width = mean(width), avg_merus = mean(raw_merus), 
            width_SD = sd(width), merus_SD = sd(raw_merus))
```

#### Summary by legal status, mean and standard deviations(SD)
```{r legal or not, echo=FALSE}

# average merus length by legal or not
dat %>% 
  group_by(legal) %>% 
  summarise(n = n(), avg_width = mean(width), avg_merus = mean(raw_merus), 
            width_SD = sd(width), merus_SD = sd(raw_merus))

```


## Linear relationship by legal status
```{r linear by group, echo = FALSE}
ggplot(dat, aes(width, raw_merus, colour = legal)) +
         geom_point(size =2) +
         geom_smooth(method =lm)


```
